Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-20 tex/draft.tex
7/19 update

\doublespacing

\section{The Standard Model}

\section{Naturalness and the Higgs mass}
    The naturalness criterion takes several forms, some of which feel way more like bullshit than others. Here are the formulations I can think of off the top of my head: (1) all parameters in a model should be of order 1, (2) Loop-level contributions should not be larger than tree-level contributions, (3) any small parameter should be protected by a custodial symmetry, and (4) fine-tuning is bad. To my ear, (1) and (2) sound dumb and (3) and (4) might be compelling with the right context.

    Let's look at (3). In the SM, the parameter of interest is the Higgs mass, which unlike the fermions and gauge bosons, has no custodial symmetry to hold it down. The fermions masses are protected by a U(1)xU(1) chiral symmetry that would exist if the fermions were massless, and the vector bosons are protected by gauge symmetry (*how does this one work? Would gauge symmetry be enlarged if vector bosons were massless?). The Higgs, a fundamental scalar, has no such symmetry. Looking at the 1-loop corrections to the Higgs propagator, we see a quadratic dependence on the cutoff scale, which implies that a Higgs mass well below the energy scale of new physics is unnatural.

    Now look at (4), again in the context of the Higgs mass. If mH depends quadratically on the cutoff, we expect it's mass to be fucking huge, like comparable to the cutoff scale. We know experimentally that its mass is 125.1 GeV, so it's bare mass (*or is it a coupling constant of some kind? I need to be extremely clear on this) must exactly cancel the cutoff-dependent contribution. 'Whatever', you might say, 'things cancel each other out sometimes'. In this case, however, the values need to be tuned to the tune of 1 part in 10\^30 or some crazy shit like that. We have no good explanation for why these two numbers would match so perfectly, so it seems incredibly suspicious that it would just happen by chance.

    There are other unnatural quantities that people worry about (even though they're less important to my world). I know of two off the top of my head: (1) whatever the parameter of interest is in the strong CP problem and (2) the cosmological consant. The strong CP problem is something like "some parameter in QCD stuff implies maximal CP violation, but we don't see any CP violation in strong interactions. The cosmological constant is way the fuck too small compared to naive expectations. Maybe like 120 orders of magnitude too small? I don't know of any reasonable solutions to the cosmological constant problem, and it seems like such a giant crazy problem that people just kinda sweep it under the rug. I have now said roughly everything I know about naturalness problems outside the contet of the Higgs mass.

\section{Naturalness Beyond the Standard Model}
    There are a few proposed solutions to the SM naturalness problem. SUSY is the main one, and I'll spend the great majority of my time on it. Extra dimensions and composite Higgs models are the other two that pop up fairly often, but won't spend much time on them outside the next two sentences. Extra dimensions solves the hierarchy problem either by lowering the cutoff scale (as in large extra dimensions (ADD)) or introducing a warping factor that reduces the level of unnaturalness (warped extra dimensions (RS)). Composite Higgs models propose that the Higgs stops behaving as a fundamental scalar at some energy level, which removes the quadratic dependence.

    SUSY introduces a new symmetry in which every field can rotate freely between fermions and bosons. Thus, every correction to the Higgs mass from a fermion loop is exactly cancelled by the assoicated boson loop, and vice-versa. We know that SUSY can't be an exact symmetry because the superpartners would have already been observed with the same mass as their SM counterparts. Broken SUSY can still be natural, however, as long as it's approximately true. That is, as long as the superpartners show up at a mass scale not too far above the EW scale. Some sparticles are more important than others for correcting the higgs mass and therefore naturalness, so we can get away with some sparticles being more massive than others, which is getting more appealing as null LHC results keep rolling in. There are other tricky things you can do to work around all these null results; I'll need to learn more about them in the immediate future.

\section{Long-lived Particles}
\subsection{Standard Model}
\subsection{Beyond the Standard Model}

@@

\section{LHC Searches for Long-lived Particles}
\subsection{Overview}
\subsection{Displaced Vertices}
            LLPs with lifetimes on the order of ~1 nm (*check this) can decay inside the detector volume, but with a reconstructed vertex far from the nominal interaction point (*last 3 words true?). ATLAS performs one such search with large MET and at least one displaced vertex (DV) with large track multiplicity. The search is sensitive to long-lived gluinos with lifetimes in the 10 ps to 10 ns range.

>>>>

\newpage
\tableofcontents
\newpage
\doublespacing

\section{The Standard Model}
    The standard model of particle physics (SM) describes all known particles and forces (except gravity) and is experimentally verified to astonishing (*weak word) accuracy (*say 2nd clause better). The SM consists of 12 spin-1/2 particles, the fermions, that make up all observed matter, 5 spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces, and 1 fundamental scalar, the Higgs boson, which will be described in detail below.

    (*Rewrite this whole paragraph while looking at a book)
    The SM particles fit into multiplets of three symmetry groups, which makes the SM an SU(3)cxSU(2)LxSU(1)Y theory (*awkward). These three symmetry groups correspond to color, the charge associated with the strong nuclear force; weak isospin, the charge associated with the left-handed chiral doublet something something; and weak hypercharge, the something something something, respectively.

    The Higgs potential (*insert equation) has a nonzero vacuum expectation value (VEV), which spontaneously breaks the SU(2)LxU(1)Y symmetry to a U(1)EM symmetry. Breaking 3 of the 4 generators leaves 3 extra degrees of freedom, which ultimately become the masses of the W+, W-, and Z bosons, while the remaining degree of freedom is associated with electric charge and the massless photon. Expanding the Higgs field about its VEV and picking a gauge produces quark and charged lepton mass terms in the SM Lagrangian. In this way, spontaneous symmetry breaking of the Higgs field gives mass to all massive SM particles (*Give example mass term or work it out a little bit).

    (*rewrite this paragraph while looking at a book)
    Like the mass of all SM particles, the Higgs mass is affected by interactions with virtual particles through loop processes. As the only fundamental scalar in the theory, however, the Higgs sees corrections to its mass that depend quadratically on the cutoff scale of the SM. (*compare to dependence of fermions and gauge bosons, show a little math). To show this dependence on loop corrections, SM masses are often written as power series, where the first term is the bare (uncorrected) mass and the following terms are due to loop corrections (*how bout some eqs?). The next section will explore the concept of naturalness, a major motivating principle in beyond the SM (BSM) physics, that is directly tied to the relationship between a particle's bare mass and the terms that correct it.

\section{Naturalness and the Higgs mass}
    The naturalness criterion takes several forms. Here are a few (*cite these): (1) all parameters in a model should be of order 1, (2) Loop-level contributions should not be larger than tree-level contributions, (3) any small parameter should be protected by a custodial symmetry, and (4) fine-tuning is bad (*probably a better way to say 4). Although these statements are all trying to get at the same fundamental notion, I find (3) and (4) to be most compelling.

    (*cite guidice naturally and t'hooft naturalness)
    Let's look at (3). In the SM, the small parameter of interest is the Higgs mass, which has no custodial symmetry to hold it down. This situation contrasts that of the fermions and gauge bosons, whose masses are kept small by custodial symmetries. That is, the SM would be more symmetric (*weird wording?) if fermions and gauge bosons were massless. Specifically, massless fermions would have an extra U(1)xU(1) chiral symmetry, and massless vector bosons would have an extra gauge symmetry (*?!?!?!). The Higgs, a fundamental scalar, has no such symmetry (*repetative) so we deem its low mass to be unnatural. (*explain why fundamental scalars have no custodial symmetry? I certainly better learn why.)

    Now look at (4), again in the context of the Higgs mass. The terms correcting the Higgs mass depend quadratically on the cutoff energy scale, which must be at or below the scale where quantum gravity becomes important (MP = 1/rootG ~ 10\^18??). This creates a sticky situation. If the terms correcting the bare mass are order(10\^32)GeV and the observed Higgs mass is order(100)GeV, then the bare mass must exactly cancel the corrections to approximately 1 part in 10\^30. At present, no good evidence exists to explain why the bare mass and corrections would match so perfectly. The level of exact cancellation is known as 'fine tuning'.

    (*probably doesn't need to be in final paper, but I'd like to check my basic understanding with Jamie, et al)
    There are other unnatural quantities that people worry about in physics. I know of two off the top of my head: (1) thetaQCD in the strong CP problem (2) the cosmological consant. The strong CP problem comes from the fact that QCD technically allows CP violation even though such violation has never been observed. The degree of violation is parameterized as thetaQCD, which recasts the problem into the naturalness context of 'why is thetaQCD so small?'. The cosmological constant is experimentally ~60 to ~120 orders of magnitude smaller than the predicted value (*predicted how?). While these issues deserve serious attention, they will not be considered further here.

    (*talk about anthropic principle or other arguments against?)
    (*I got tired of reading this section and haven't edited this paragraph at all yet)
    There are a few proposed solutions to the SM naturalness problem. Supersymmetry (SUSY) is the main one, and I'll spend the great majority of my time on it. Extra dimensions and composite Higgs models are the other two that pop up fairly often, but won't spend much time on them outside the next two sentences. Extra dimensions solves the hierarchy problem either by lowering the cutoff scale (as in large extra dimensions (ADD)) or introducing a warping factor that reduces the level of unnaturalness (warped extra dimensions (RS)). Composite Higgs models propose that the Higgs stops behaving as a fundamental scalar at some energy level, which removes the quadratic dependence. (*Hidden valley, dark sector, little Higgs?)

\section{SUSY and LLPs}
    (*is there a way to move the naming stuff elsewhere or shorten it?)
    SUSY introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between fermions and bosons. In it's simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson (*and an extra Higgs field - can I gloss over that?). The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-1/2 superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles. Because fermion loops come with a relative minus sign (*there must be a better way to say that), every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. If supersymmetry were exact, the bare Higgs mass would be uncorrected to all orders in perturbation theory.

    Experimental evidence shows that SUSY must be broken, however, because no superpartners have been found, and exact SUSY requires each superpartner to have the same mass as its standard model counterpart. The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenemena. A handful of models with particular approaches to SUSY breaking that are particularly prominant or particularly relevant to naturalness or long-lived particles (LLPs) are gauge-mediated susy breaking (GMSB), anomaly-mediated SUSY breaking (AMSB), split SUSY, and R-parity violating SUSY (RPV). The details of these models will be explored as needed throughout this paper.

    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and (*and/or?) would have reduced sensitivity to common SUSY search techniques (*does 'reduced sensitivity to techniques' make sense?). Specifically, we are interested in SUSY breaking models that predict LLPs. That is, particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point, which includes everything from particles that decay at (*give ballparky number (~100um?)) to particles that propagate through the entire detector.

    In both the SM and in BSM theories, LLPs result from one of the following: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, the electron and proton are the lightest particles carrying electric charge and baryon number, respectively, which leads to their absolute stability through charge and baryon number conservation. Neutrons and muons, on the other hand, are long-lived beause the proton and electron masses leave them little phase space in which to decay (*also mention that they're both weak decays?). (*Mention neutrino, pion, kaon, whatever other bound state I might be forgetting?)

    In SUSY models, LLPs arise for the same basic reasons, and many different SUSY models meet different LLP criteria in different ways. The most common (*based on my assumptions) SUSY LLP occurs in models with conserved R-partity, which is a proposed symmetry that protects many models from predicting proton decay. In models with R-parity, the SM particles have R-parity +1 (*double check this explanation) and their superpartners have R-parity -1. Conserving overall R-parity (*specifically the product; does overall mean that?) at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is an absolutely stable LLP. The second phenomenon is the main motivation behind using missing energy as the typical SUSY seach signature.

    In models with R-parity violation (RPV), LLPs can be produced when the RPV terms have small coupling constants, which leads to long-lived, but not perfectly stable, LSPs (*any way non-LSPs could be long-lived for this reason too?). Luckily, RPV models can still avoid proton decay by only allowing terms that violate either baryon number or lepton number but not both. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs(*do I need to explain GMSB here?). (*cite dimopoulos low energy)

    LLPs from highly-virtual intermediate particles can occur in models such as split (or mini-split) SUSY, where the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. (*cite kilian) (*mention that split throws out naturalness motivation?)

    SUSY models with degenerate sparticle spectra can have increased sparticle lifetimes from the reduced phase space for decays. In some AMSB models (*be more specific), for example, the zino and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino's lifetime, then, is increased by the lack of phase space in which to decay. (*cite R\&S out of this world)
    (*merge last 2 paragraphs under 'mass spectra lead to LLP' concept?)
    (*never said anything about hidden valley or dark sector stuff. Never said the word R-hadron. Never said stealth or displaced SUSY. Should I?)

@@

\section{LHC Searches for Long-lived Particles}
\subsection{Overview}
    LLPs leave distinct signatures that differentiate LLP signals from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were generally designed to detect and reconstruct stable particles from prompt decays, so LLP analyses often need to veer off the traditional analysis path in terms of object reconstruction (*really reconstruction? Maybe object definition?), background estimation, and which data tier to use (*can I say this more generally?). This paper will focus on four complementary LLP signatures: displaced vertices, disappearing tracks, stopped particles, and heavy stable charged particles (HSCPs)(*is HSCP a signature? say TOF / dE/dx instead?).

    Displaced vertices and disappearing tracks both look for particles that are unstable on detector timescales (*quantify?). Displaced vertices searches are sensitive to LLPs that decay to visible (*detectable? some other word?) particles in the inner volume (*specify) of the detector and leave tracks that can be traced back to a vertex that is sufficiently displaced from the nominal interaction point. Disappearing tracks searches are sensitive to LLPs that decay to invisible (*neutral? some other word?) particles along with particles that are too soft to be reconstructed. Together, these two types of analyses cover something something parameter space of LLPs that decay on detector timescales.

    HSCP and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. HSCP searches use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles, which normally (*quantify) have beta > 0.9. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomena that could be possible with long-enough LLP lifetimes and small initial kinetic energy. These analyses look (*repetative) for calorimeter energy deposits or hits in the muon system (*too specific for LLP analyses intro?) that are out of time with proton bunch crossings that could be due to stopped particles decaying.

    These four approaches cover a wide range of LLP lifetimes that would be missed in traditional searches. The next section will look  at a recent example of each analysis type from either ATLAS or CMS.
\subsection{Displaced Vertices}
    LLPs with lifetimes on the order of ~1 nm (*check this) can decay inside the detector volume, but with a reconstructed vertex far from the nominal interaction point (*last 3 words true?). ATLAS performs one such search with large MET and at least one displaced vertex (DV) with large track multiplicity. The search is sensitive to long-lived gluinos with lifetimes in the 10 ps to 10 ns range.
